\LARGE \bf Planning with Attitude

### 2020-09-09 paper/main.tex
Move IEEE version to main.tex

% Formatting
\usepackage{xcolor}
\usepackage{bm}  % for bold symbols 
\usepackage{booktabs}  % better tables
\usepackage{pifont}  % for x mark
\usepackage{subcaption}
\captionsetup{compatibility=false}
\usepackage{wrapfig}
\usepackage{graphicx}

@@

%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ BEGIN DOCUMENT ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~%
\begin{document}
\mainmatter              % start of a contribution

\title{Planning and Control with Attitude}
\author{Brian Jackson \and Zachary Manchester}

\authorrunning{Jackson and Manchester}
\institute{Stanford University, 496 Lomita Mall, Stanford, CA, USA,\\
\email{\{bjack205,zacm\}@stanford.edu}}

\maketitle

\begin{abstract}
Planning and controlling trajectories for floating-base robotic systems that experience large attitude changes is challenging due to the nontrivial group structure of 3D rotations. 
This paper introduces an accessible and unified approach for tracking control and optimization-based planning on the space of rotations based on vector calculus and linear algebra. The methodology is used to derive an extension of the Linear-Quadratic Regulator (LQR) to systems with arbitrary attitudes, which we call Multiplicative LQR (MLQR). 
We compare MLQR to a state-of-the-art geometric tracking controller designed specifically for quadrotors and demonstrate superior tracking and robustness. 
An iterative variant of MLQR is compared to existing methods on useveral constrained trajectory optimization problems, again demonstrating improved runtime performance. %to optimize trajectories for a variety of robotic systems, we provide benchmark comparisons between several of the most common attitude representations used for motion planning and control and find that the combination of unit quaternion state variables with Rodrigues parameters to represent attitude errors provides an excellent combination of performance and algorithmic simplicity. 
\keywords{motion planning and control, quaternions, optimal control, linear quadratic regulator}
\end{abstract}

\section{Introduction}
    Many useful robotic systems---including quadrotors, airplanes, satellites, autonomous underwater vehicles, and quadrupeds---can perform arbitrarily large three-dimensional translations and rotations as part of their normal operation. 
    While representing translations is straightforward and intuitive, effectively representing the nontrivial group structure of 3D rotations has been a topic of study for many decades. Although we can intuitively deduce that rotations are three-dimensional, a globally non-singular three-parameter representation of the space of rotations does not exist \cite{stuelpnagel1964parametrization}. As a result, when parameterizing rotations, we must either a) pick a three-parameter representation and deal with discontinuities, or b) pick a higher-dimensional representation and deal with constraints between the parameters. While simply representing attitude is nontrivial, generating and tracking motion plans for floating-base systems is an even more challenging problem.

    Early work on control problems involving the rotation group dates back to the 1970s, with extensions of linear control theory to spheres \cite{Brockett1973} and $SO(3)$ \cite{Baillieul1978}. Effective attitude tracking controllers have been developed for satellites \cite{wie1985quaternion}, quadrotors \cite{Fresk2013,Liu2015,lee2010geometric,Johnson2005,watterson2020control,mellinger2011minimum}, and a 3D inverted pendulum \cite{Chaturvedi2009}  using various methods for calculating three-parameter attitude errors.

    More recently, these ideas have been extended to trajectory generation \cite{Zefran1998}, sample-based motion planning \cite{Zefran1999,Kuffner2004}, and optimal control. Approaches to optimal control on attitude problems include analytical methods applied to satellites \cite{Spindler1998}, discrete mechanics \cite{Kobilarov2011,Kobilarov2014,Lee2008}, a combination of sampling-based planning and constrained trajectory optimization for satellite formations \cite{Garcia2005, Aoude2008}, projection operators \cite{Saccon2013}, or more general theory for optimization on manifolds \cites{watterson2018trajectory}. Nearly all of these methods rely heavily on principles from differential geometry and Lie group theory; however, despite these works, many recent papers in the robotics community continue to apply traditional methods for motion planning and control with no regard for the group structure of rigid body motion \cite{Alothman2016,deCrousaz2015,Williams2017,Geisert2016}.

    In this paper, we make a departure from previous approaches to geometric planning and control that rely heavily on ideas and notation from differential geometry, and instead use only basic mathematical tools from linear algebra and calculus that should be familiar to most roboticists. Similar to \cite{Mandic2011,Xu2016}, in Sec. \ref{sec:Quaternion_Calculus} we introduce a quaternion differential calculus, but take a significantly simpler and more general approach that unifies both planning and control, enabling straight-forward adaptation of existing algorithms to systems with quaternion states. To make this concrete, in Sec. \ref{sec:MLQR} we apply our approach to derive an extension to the linear-quadratic regulator (LQR) that we call multiplicative LQR (MLQR), which is the control dual to the multiplicative extended Kalman filter (MEKF) from the state estimation literature \cite{markley2014fundamentals}. In Sec. \ref{sec:experiments} we provide several simulation results demonstrating the application of our method to tracking control and constrained trajectory optimization.
    In summary, our contributions include:
    %\vspace{-10pt}
    \begin{itemize}
        \item A unified approach to quaternion differential calculus entirely based on standard vector calculus and linear algebraic operations
        \item Derivation of multiplicative LQR, an adaptation of LQR to the control of systems with quaternion states
        \item A set of benchmark motion-planning and trajectory-tracking problems in which we compare our approach to existing methods commonly used in the robotics community to account for 3D rotations
    \end{itemize}

\section{Background}
    We begin by defining some useful conventions and notation. Attitude is defined as the rotation from the robot's body frame to a global inertial frame. We also define gradients to be row vectors, that is, for $f(x) : \R^n \to \R$, $\pdv{f}{x} \in \R^{1\times n}$.

@@

    \begin{figure}
        \centering
        % \includegraphics[height=6cm]{figures/tangent_plane.png}
        \includegraphics[height=5cm]{figures/tangent_plane.tikz}
        \caption{When linearizing about a point $q$ on an sphere $\mathbb{S}^{n-1}$ in n-dimensional space, the tangent space $T$ is a plane living in $\R^{n-1}$, illustrated here with $n=3$. Therefore, when linearizing about a unit quaternion $q \in \Q$, the space of differential rotations lives in $\R^3$.}
        \label{fig:tangent_plane}
    \end{figure}

        Derivatives consider the effect an infinitesimal perturbation to the input has on an infinitesimal perturbation to the output. For vector spaces, the composition of the perturbation with the nominal value is simple addition and the infinitesimal perturbation lives in the same space as the original vector. For unit quaternions, however, neither of these are true; instead, they compose according to \eqref{eq:quat_mult}, and infinitesimal unit quaternions are (to first order) confined to a 3-dimensional plane tangent to $\Q$ (see Fig. \ref{fig:tangent_plane}).
        %, and are related through a non-linear mapping $\varphi : \R^3 \to \Q$ )
        The fact that differential unit quaternions are three-dimensional should make intuitive sense: Rotations are inherently three-dimensional and differential rotations should live in the same space as angular velocity, i.e. $\R^3$.

        There are many possible three-parameter representations for small rotations in the literature. Many authors use the exponential map \cite{Baillieul1978,Zefran1998,Lee2008,Saccon2013,Sola2017,Fan2016,watterson2018trajectory}, while others have used the Cayley map (also known as Rodrigues parameters) \cite{Kobilarov2011,Kobilarov2014}, Modified Rodrigues Parameters (MRPs) \cite{Terzakis2018}, or the vector part of the quaternion \cite{Fresk2013}. 
        We choose Rodrigues parameters \cite{markley2014fundamentals} because they are computationally efficient and do not inherit the sign ambiguity associated with unit quaternions. The mapping between a vector of Rodrigues parameters $\phi \in \R^3$ and a unit quaternion $q$ is known as the Cayley map:
        \begin{equation} \label{eq:cayley}
            q = \varphi(\phi) = \frac{1}{\sqrt{1 + \norm{\phi}^2}} \begin{bmatrix} 1 \\ \phi \end{bmatrix}.
        \end{equation}
        We will also make use of the inverse Cayley map:
        \begin{equation}
            \phi = \varphi^{-1}(q) = \frac{q_v}{q_s}.
        \end{equation}

    \subsection{Jacobian of Vector-Valued Functions}
        When taking derivatives with respect to quaternions, we must take into account both the composition rule and the nonlinear
        mapping between the space of unit quaternions and our chosen three-parameter error representation. 
        % Before demonstrating how this is done, let's consider the real-valued case we should already be familiar with. For some function $y = r(x) : \R^n \to \R^p$ with a differential value $\delta x \in \R^n$ applied to the input:
        % \begin{align}
        %     y + \delta y = r(x+\delta x) &\approx r(x) + \nabla r(x) \delta x,
        %     % &= \lim_{\delta x \to 0} x + \pdv{r}{x} \pdv{r + \delta x}{\delta x} = r + \pdv{r}{x} 
        % \end{align}
        % we calculate $\nabla r$ by taking the derivative with respect $\delta x$, as $\delta x$ approaches zero:
        % \begin{equation}
        %     \nabla r(x) = \lim_{\delta x \to 0} \pdv{r}{x} \pdv{(x + \delta x)}{\delta x} = \lim_{\delta x \to 0} \pdv{r}{x} I = \pdv{r}{x}
        % \end{equation}
        Let $\phi \in \R^3$ be a differential rotation applied to a function with quaternion inputs $y = h(q): \Q \to \R^p$, such that
        \begin{equation} \label{eq:vector_function}
            y + \delta y = h(L(q) \varphi(\phi)) \approx h(q) +  \nabla h(q) \phi.
        \end{equation}
        We can calculate the Jacobian $\nabla h(q) \in \R^{p \times 3}$ by differentiating \eqref{eq:vector_function} with respect to $\phi$, evaluated at $\phi = 0$:
        \begin{equation} \label{eq:quat_gradient}
            \nabla h(q) = \pdv{h}{q} L(q) H := \pdv{h}{q} G(q) = \pdv{h}{q} \begin{bmatrix} -q_v^T \\ sI_3 + \skewmat{q_v} \end{bmatrix}
        \end{equation}
        where $G(q) \in \R^{4 \times 3}$ is the \textit{attitude Jacobian}, which essentially becomes a ``conversion factor'' allowing us to apply results from standard vector calculus to the space of unit quaternions. This form is particularly useful in practice since $\pdv*{h}{q} \in \R^{p \times 4}$ can be obtained using finite difference or automatic differentiation.
        As an aside, although we have used Rodrigues parameters, $G(q)$ is actually the same (up to a constant scaling factor) for any choice of three-parameter attitude representation.

    \subsection{Hessian of Scalar-Valued Functions}
	    If the output of $h$ is a scalar ($p = 1$), then we can find its Hessian by taking the Jacobian of \eqref{eq:quat_gradient} with respect to $\phi$ using the product rule, again evaluated at $\phi = 0$:
	   % \begin{equation} \label{eq:hess1}
	   %     \nabla^2 h(q) = \lim_{\phi\to0} \left(L(q) \pdv{\varphi}{q} \right)^T \pdv[2]{h}{q} \left(L(q) \pdv{\varphi}{\phi} \right)+ \pdv{\phi}  \left(G(L(q) \varphi(\phi))^T \pdv{h}{q}^T \right) .
	   % \end{equation}
	   % Expanding the second term in \eqref{eq:hess1} and simplifying results in:
	    \begin{equation} \label{eq:quat_hessian}
            \nabla^2 h(q) = G(q)^T \pdv[2]{h}{q} G(q) + I_3 \pdv{h}{q}q,
	    \end{equation}
	    where the second term comes from the second derivative of $\varphi(\phi)$. Similar to $G(q)$, this ends up being the same (up to a scaling factor) for any choice of three-parameter attitude representation.

@@

    % We now wish to find a sequence of controls that minimize a quadratic cost function penalizing deviations from the reference trajectory and large control values, while obeying our system dynamics. This is succinctly stated by the following optimization problem:
    % \begin{mini}[2]
    % 	{x_{0:N},u_{0:N-1}}{\half x_N^T W_N x_N + \half\sum_{k=0}^{N-1} x_k^T W_k x_k + u_k^T R_k u_k }{}{}
    % 	\addConstraint{x_{k+1}}{=A_k x_k + B_k u_k}
    % 	\addConstraint{x_0}{=x_\text{init}}
    % 	\label{opt:discrete_LQR},
    % \end{mini}
    % where we have dropped the $\delta$ leading the states and controls to reduce visual clutter.

    % We now define the \textit{value function} $V_k(x)$ which is simply the cost beginning at time step $k$, such that $V_0(x)$ is the total cost, and $V_N(x) = \half x_N^T W_N x_N$ is the terminal cost. Using the Bellman equation and the principle of optimality, we can define it recursively: %, along with the \textit{action-value function} $Q(x_k,u_k)$:
    % \begin{equation} \label{eq:value_function}
    %     V_k(x) = \min_{u_k} \half x_k^T W_k x_k + \half u_k^T R_k u_k + V_{k+1}(f(x_k,u_k))
    % \end{equation}
    % % \begin{equation} \label{eq:action_value_function}
    % %     V_k(x) = \min_{u_k} Q(x_k,u_k)
    % % \end{equation}
    % By assuming $V_k(x) = \half x_k^T P_k x_k$ for some symmetric positive-definite matrix $P$, we can solve \eqref{eq:value_function} in terms of $P_{k+1}$ using the first-order necessary condition:
    % \begin{equation}
    %     \pdv{Q}{u} = R_k u_k + B^T P_{k+1}(A_k x_k + B_k u_k) = 0
    % \end{equation}
    % Solving for $u_k$ we find the optimal feedback policy: 
    % \begin{equation} \label{eq:LQR_gain}
    %     u_k^* = -(R_k + B_k^T P_{k+1} B_k)^{-1} B_k^T P_{k+1} A_k x_k \equiv -K_k x_k
    % \end{equation}
    % % where $Q_{uu} = \pdv*[2]{Q}{u}$ and $Q_{ux} = \pdv*{Q}{u}{x}$.

    % With an optimal feedback policy, we now substitute \eqref{eq:LQR_gain} into \eqref{eq:value_function}, and after simplifying the the result to a quadratic expression in $x_k$, obtain the recurrence relationship:
    % \begin{equation}
    %      P_k = W_k +  A_k^T P_{k+1} A_k - A_k^T P_{k+1}^T B_k (R_k + B_k^T P_{k-1} B_k)^{-1} B_k^T P_{k+1} A_k.
    % \end{equation}
    % Iterating backwards in time defines the control gains $K \in \R^{m \times 12}$ for each time step. These gains are cached in the controller and looked up online.

    % When running the controller online, we compare our current state estimate $\hat{x}$ to our reference state $\bar{x}$ and compute our state differential $\delta x$ using the nonlinear function \eqref{eq:state_error}. The new control value is then computed as

@@

        % While the core algorithm doesn't change in practice, sometimes the approach used to solve the problem has to be modified, especially with cost functions. Since many cost function penalize distance from some desired state, there are better ways of doing this than naively subtracting two quaternions. The following section describes a few cost functions that are better-suited for this type of task.

        % \begin{algorithm}
        % \begin{algorithmic}[1]
        % \caption{iMLQR} \label{alg:iMLQR}
        % \Function{ForwardPass}{$\traj[N-1]{u}$} 
        %     \State $x,u \leftarrow $ simulate the system forward 
        %     \While{not converged}
        %         \State {\color{blue} $W,R,w,r \leftarrow$ Quadratic expansion \eqref{eq:quat_gradient}, \eqref{eq:quat_hessian}}
        %         \State {\color{blue} $A,B \leftarrow$ Linearize dynamics \eqref{eq:linearized_dynamics}}
        %         \State $K,d,\Delta V \leftarrow$ \textproc{LQR}$(W,R,w,r,A,B)$
        %         \State {\color{blue} $x,u \leftarrow$ \textproc{ForwardPass}$(\traj{x},\traj{u},K,d,\Delta V)$}
        %     \EndWhile\\
        %     \Return $X,U,J$
        % \EndFunction
        % \end{algorithmic}
        % \end{algorithm}

        % \begin{algorithm}
        % \begin{algorithmic}[1]
        % \caption{Forward Pass} \label{alg:FP}
        % \Function{ForwardPass}{$X,U,K,d,\Delta V,J$} 
        %     \State Initialize $\bar{x}_0 = x_0$, $\alpha = 1$, $J^{\small{-}} \leftarrow J$
        %     \For{k=0:1:N-1}
        %         \State {\color{blue} $\delta x \leftarrow$ Using \eqref{eq:state_error}}
        %         \State $\bar{u}_k = u_k + K_k \delta x + \alpha d_k$
        %         \State $\bar{x}_{k+1} \leftarrow f(\bar{x}_k, \bar{u}_k)$
        %     \EndFor
        %     \State $J \leftarrow$ Calculate cost using $\traj{x},\traj{u}$ 
        %     \If{$J$ {\color{blue} satisfies line search conditions}}
        %         \State $x_{0:N} \leftarrow \bar{x}_{0:N}, u_{0:N-1} \leftarrow \bar{u}_{0:N-1}$
        %     \Else
        %         \State Reduce $\alpha$ and go to line 3
        %     \EndIf\\
        %     \Return $X,U,J$
        % \EndFunction
        % \end{algorithmic}
        % \end{algorithm}

@@

        \subsubsection{Error Quadratic}
            Rather than simple subtraction, we can use a quadratic function on the three-parameter error state \eqref{eq:state_error}:
            \begin{equation} \label{eq:error_quadratic}
                J_\text{err} = \half \phi^T Q \phi = \half \left(\varphi^{-1}(\delta q)\right)^T Q \left(\varphi^{-1}(\delta q)\right).
            \end{equation}
            where $\delta q = L(q_g)^T q$, and $\phi = \varphi{\delta q}$. The gradient and Hessian of \eqref{eq:error_quadratic} are 
            \begin{equation}
                \nabla J_\text{err }= \phi^T Q  D(\delta q)  G(\delta q)
            \end{equation}
            \begin{equation}
                \begin{aligned}
                     \nabla^2 J_\text{err} &= 
                        G(\delta q)^T \! \left(
                        D(\delta q)^T Q D(\delta q) + \nabla D \right) G(\delta q) 
                      + I_3 (\phi^T Q  D(\delta q)) \delta q %\\
                    %  &+ G(\delta q)^T \nabla D \left(\delta q, Q \phi \right) G(\delta q) \\
                \end{aligned}
            \end{equation}
            where, for the Cayley map,
            \begin{equation}
                D(q) = \pdv{\varphi^{-1}}{q} = -\frac{1}{q_s^2}\begin{bmatrix}
                    q_v \;\; & -\frac{1}{q_s} I_3
                \end{bmatrix}
            \end{equation}
            \begin{equation}
                \nabla D = \pdv{q}(D(q)^T Q \phi) 
                = -\frac{1}{q_s^2} \begin{bmatrix} -2 \frac{q_v}{q_s}^T Q \phi \;\;  & \phi^T Q \\
                                   Q \phi \;\; & 0 \\
                \end{bmatrix}.
            \end{equation}

            % where $\delta q \in \R^4 = L(q_g)^T q$, $\psi \in \R^3 = \varphi(\delta q)$, and $\nabla \varphi^{-1}(q) \in \R^{3 \times 4} \equiv \pdv*{\varphi^{-1}}{q}$.

            % \begin{equation}
            %     \nabla_{x} = \psi^T Q \pdv{\varphi^{-1}}{q} L(\delta q) \pdv{\varphi}{\phi}
            % \end{equation}

            % \noindent We can obtain the Hessian through the chain rule:
            % \begin{equation}
            %     \begin{aligned}
            %          \nabla^2_{xx} &= \pdv{\varphi}{\phi}^T L(\delta q)^T \left(\pdv{\varphi^{-1}}{q}\right)^T Q \left(\pdv{\varphi^{-1}}{q}\right) L(\delta q) \pdv{\varphi}{\phi} \\
            %          &+ \pdv{\varphi}{\phi}^T L(\delta q)^T 
            %          \nabla^2 \varphi^{-1} \left(\delta q, Q \psi \right) L(\delta q) \pdv{\varphi}{\phi} \\
            %          &+ \nabla^2 \varphi \left(\delta q, L(\delta q)^T \pdv{\varphi^{-1}}{q} Q \psi \right)
            %     \end{aligned}
            % \end{equation}
            % Evaluating at $\phi = 0$ we get the gradient

@@

\section{Experiments} \label{sec:experiments}
    Code code for all experiments is available on GitHub at \url{\texttt{https://github.com/RoboticExplorationLab/PlanningWithAttitude}}.

@@

        \begin{figure}
            \centering
            \includegraphics[height=4cm, width=8cm]{figures/timing_chart.tikz}
            \caption{Timing results for the airplane barrell roll, quadrotor flip, and quadrotor zig-zag examples using iLQR with roll-pitch-yaw Euler angles (RPY), quaternions, and iMLQR. The timing result for the Quadrotor flip with Euler angles is omitted since it failed to converge.}
            \label{fig:timing_chart}
        \end{figure}

        \subsubsection{Quadrotor Zig-Zag}
            \begin{figure}[t]
                \centering
                \begin{subfigure}[t]{0.55\textwidth}
                    \includegraphics[height=5.2cm]{figures/zig_zag_tracking.png}
                    \caption{Zig-zag trajectory. The black quadrotors indicate the reference trajectory generated by iMLQR. The yellow quadrotors track the reference trajectory using time-varying MLQR, while the red uses the HFCA controller from \cite{watterson2020control}.}
                    \label{fig:zig-zag}
                \end{subfigure}
                ~
                \begin{subfigure}[t]{0.40\textwidth}
                    \centering
                    \includegraphics[height=5.2cm,trim={0 3cm 0 0},clip]{figures/quad_flip.png}
                    \caption{Snapshots of the quadrotor flip trajectory. The red-colored quadrotors represent the state near t=0 s and the green-colored quadrotors represent the state near t=5.0 s}
                    \label{fig:quad_flip}
                \end{subfigure}
                \caption{Examples applying MLQR to trajectory optimization on quadrotors}
            \end{figure}

>>>>

% Formatting
\usepackage{xcolor}
\usepackage{bm}  % for bold symbols 
\usepackage{booktabs}  % better tables
\usepackage{pifont}  % for x mark
\usepackage{graphicx}
\usepackage{hyperref}

@@

\title{\LARGE \bf
Planning with Attitude
}

\author{Brian Jackson$^1$ and Zachary Manchester$^1$%
    \thanks{
        $^1$Robotics Institute, 
        Carnegie Mellon University, 
        5000 Forbes Ave, Pittsburgh, PA, USA
    }
}

\begin{document}
\maketitle

\begin{abstract}
Planning and controlling trajectories for floating-base robotic systems that experience
large attitude changes is challenging due to the nontrivial group structure of 3D
rotations. This paper introduces an accessible and unified approach for tracking control
and optimization-based planning on the space of rotations based on vector calculus and
linear algebra. The methodology is used to derive an extension of the Linear-Quadratic
Regulator (LQR) to systems with arbitrary attitudes, which we call Multiplicative LQR
(MLQR). We compare MLQR to a state-of-the-art geometric tracking controller designed
specifically for quadrotors and demonstrate superior tracking and robustness. An
iterative variant of MLQR is compared to existing methods on several constrained
trajectory optimization problems, again demonstrating improved runtime performance. 
\keywords{
    motion planning and control, quaternions, optimal control, linear quadratic regulator
}
\end{abstract}

\section{INTRODUCTION}
    Many useful robotic systems---including quadrotors, airplanes, satellites, autonomous 
    underwater vehicles, and quadrupeds---can perform arbitrarily large three-dimensional 
    translations and rotations as part of their normal operation. 
    While representing translations is straightforward and intuitive, effectively 
    representing the nontrivial group structure of 3D rotations has been a topic of study 
    for many decades. Although we can intuitively deduce that rotations are three-dimensional, 
    a globally non-singular three-parameter representation of the space of rotations does 
    not exist \cite{stuelpnagel1964parametrization}. As a result, when parameterizing rotations, 
    we must either a) pick a three-parameter representation and deal with discontinuities, 
    or b) pick a higher-dimensional representation and deal with constraints between the 
    parameters. While simply representing attitude is nontrivial, generating and tracking 
    motion plans for floating-base systems is an even more challenging problem.

    Early work on control problems involving the rotation group dates back to the 1970s, 
    with extensions of linear control theory to spheres \cite{Brockett1973} and $SO(3)$ 
    \cite{Baillieul1978}. Effective attitude tracking controllers have been developed for 
    satellites \cite{wie1985quaternion}, quadrotors \cite{Fresk2013,Liu2015,lee2010geometric,
    Johnson2005,watterson2020control,mellinger2011minimum}, and a 3D inverted pendulum 
    \cite{Chaturvedi2009}  using various methods for calculating three-parameter attitude 
    errors.

    More recently, these ideas have been extended to trajectory generation \cite{Zefran1998}, 
    sample-based motion planning \cite{Zefran1999,Kuffner2004}, and optimal control. 
    Approaches to optimal control on attitude problems include analytical methods applied to
    satellites \cite{Spindler1998}, discrete mechanics \cite{Kobilarov2011,Kobilarov2014,
    Lee2008}, a combination of sampling-based planning and constrained trajectory 
    optimization for satellite formations \cite{Garcia2005, Aoude2008}, projection operators 
    \cite{Saccon2013}, or more general theory for optimization on manifolds 
    \cites{watterson2018trajectory}. 
    Nearly all of these methods rely heavily on principles from differential geometry and 
    Lie group theory; however, despite these works, many recent papers in the robotics 
    community continue to apply traditional methods for motion planning and control with no 
    regard for the group structure of rigid body motion \cite{Alothman2016,deCrousaz2015,
    Williams2017,Geisert2016}.

    In this paper, we make a departure from previous approaches to geometric planning and 
    control that rely heavily on ideas and notation from differential geometry, 
    and instead use only basic mathematical tools from linear algebra and calculus that 
    should be familiar to most roboticists. 
    Similar to \cite{Mandic2011,Xu2016}, in Sec. \ref{sec:Quaternion_Calculus} we introduce 
    a quaternion differential calculus, but take a significantly simpler and more general 
    approach that unifies both planning and control, enabling straight-forward adaptation of 
    existing algorithms to systems with quaternion states. 
    To make this concrete, in Sec. \ref{sec:MLQR} we apply our approach to derive an 
    extension to the linear-quadratic regulator (LQR) that we call multiplicative LQR (MLQR), 
    which is the control dual to the multiplicative extended Kalman filter (MEKF) from the 
    state estimation literature \cite{markley2014fundamentals}. 
    In Sec. \ref{sec:experiments} we provide several simulation results demonstrating the 
    application of our method to tracking control and constrained trajectory optimization.
    In summary, our contributions include:

    \begin{itemize}
        \item A unified approach to quaternion differential calculus entirely based on standard vector calculus and linear algebraic operations
        \item Derivation of multiplicative LQR, an adaptation of LQR to the control of systems with quaternion states
        \item A set of benchmark motion-planning and trajectory-tracking problems in which we compare our approach to existing methods commonly used in the robotics community to account for 3D rotations
    \end{itemize}

\section{Background}

    We begin by defining some useful conventions and notation. 
    Attitude is defined as the rotation from the robot's body frame to a global inertial 
        frame. 
    We also define gradients to be row vectors, that is, for 
        $f(x) : \R^n \to \R$, $\pdv{f}{x} \in \R^{1\times n}$.

@@

    \begin{figure}
        \centering
        \includegraphics[height=5cm]{figures/tangent_plane.tikz}
        \caption{
            When linearizing about a point $q$ on an sphere $\mathbb{S}^{n-1}$ in 
            n-dimensional space, the tangent space $T$ is a plane living in $\R^{n-1}$, 
            illustrated here with $n=3$. Therefore, when linearizing about a unit 
            quaternion $q \in \Q$, the space of differential rotations lives in $\R^3$.
        }
        \label{fig:tangent_plane}
    \end{figure}

        Derivatives consider the effect an infinitesimal perturbation to the input has on
        an infinitesimal perturbation to the output. For vector spaces, the composition
        of the perturbation with the nominal value is simple addition and the
        infinitesimal perturbation lives in the same space as the original vector. For
        unit quaternions, however, neither of these are true; instead, they compose
        according to \eqref{eq:quat_mult}, and infinitesimal unit quaternions are (to
        first order) confined to a 3-dimensional plane tangent to $\Q$ (see Fig.
        \ref{fig:tangent_plane}).

        The fact that differential unit quaternions are three-dimensional should make
        intuitive sense: Rotations are inherently three-dimensional and differential
        rotations should live in the same space as angular velocity, i.e. $\R^3$.

        There are many possible three-parameter representations for small rotations in
        the literature. Many authors use the exponential map \cite{Baillieul1978,
        Zefran1998, Lee2008, Saccon2013, Sola2017, Fan2016, watterson2018trajectory},
        while others have used the Cayley map (also known as Rodrigues parameters)
        \cite{Kobilarov2011, Kobilarov2014}, Modified Rodrigues Parameters (MRPs)
        \cite{Terzakis2018}, or the vector part of the quaternion \cite{Fresk2013}.
        We choose Rodrigues parameters \cite{markley2014fundamentals} because they are
        computationally efficient and do not inherit the sign ambiguity associated with
        unit quaternions. The mapping between a vector of Rodrigues parameters $\phi \in
        \R^3$ and a unit quaternion $q$ is known as the Cayley map: \begin{equation}
        \label{eq:cayley}
            q = \varphi(\phi) = \frac{1}{\sqrt{1 + \norm{\phi}^2}} \begin{bmatrix} 1 \\ \phi \end{bmatrix}.
        \end{equation}
        We will also make use of the inverse Cayley map:
        \begin{equation}
            \phi = \varphi^{-1}(q) = \frac{q_v}{q_s}.
        \end{equation}

    \subsection{Jacobian of Vector-Valued Functions}
        When taking derivatives with respect to quaternions, we must take into account
        both the composition rule and the nonlinear mapping between the space of unit
        quaternions and our chosen three-parameter error representation.

        Let $\phi \in \R^3$ be a differential rotation applied to a function with
        quaternion inputs $y = h(q): \Q \to \R^p$, such that
        \begin{equation} \label{eq:vector_function}
            y + \delta y = h(L(q) \varphi(\phi)) \approx h(q) +  \nabla h(q) \phi.
        \end{equation}
        We can calculate the Jacobian $\nabla h(q) \in \R^{p \times 3}$ by
        differentiating \eqref{eq:vector_function} with respect to $\phi$, evaluated at
        $\phi = 0$:
        \begin{equation} \label{eq:quat_gradient}
            \nabla h(q) = \pdv{h}{q} L(q) H := \pdv{h}{q} G(q) 
                        = \pdv{h}{q} \begin{bmatrix} 
                            -q_v^T \\ 
                            sI_3 + \skewmat{q_v}
                        \end{bmatrix}
        \end{equation}
        where $G(q) \in \R^{4 \times 3}$ is the \textit{attitude Jacobian}, which
        essentially becomes a ``conversion factor'' allowing us to apply results from
        standard vector calculus to the space of unit quaternions. This form is
        particularly useful in practice since $\pdv*{h}{q} \in \R^{p \times 4}$ can be
        obtained using finite difference or automatic differentiation.
        As an aside, although we have used Rodrigues parameters, $G(q)$ is actually the
        same (up to a constant scaling factor) for any choice of three-parameter attitude
        representation.

    \subsection{Hessian of Scalar-Valued Functions}
	    If the output of $h$ is a scalar ($p = 1$), then we can find its Hessian by
	    taking the Jacobian of \eqref{eq:quat_gradient} with respect to $\phi$ using the
        product rule, again evaluated at $\phi = 0$:

	    \begin{equation} \label{eq:quat_hessian}
            \nabla^2 h(q) = G(q)^T \pdv[2]{h}{q} G(q) + I_3 \pdv{h}{q}q,
	    \end{equation}
	    where the second term comes from the second derivative of $\varphi(\phi)$.
	    Similar to $G(q)$, this ends up being the same (up to a scaling factor) for any
        choice of three-parameter attitude representation.

@@

@@

@@

        \subsubsection{Error Quadratic}
            Rather than simple subtraction, we can use a quadratic function on the
            three-parameter error state \eqref{eq:state_error}:
            \begin{equation} \label{eq:error_quadratic}
                J_\text{err} = \half \phi^T Q \phi 
                = \half \left(\varphi^{-1}(\delta q)\right)^T Q 
                        \left(\varphi^{-1}(\delta q)\right).
            \end{equation}
            where $\delta q = L(q_g)^T q$, and $\phi = \varphi{\delta q}$. The gradient
            and Hessian of \eqref{eq:error_quadratic} are
            \begin{equation}
                \nabla J_\text{err }= \phi^T Q  D(\delta q)  G(\delta q)
            \end{equation}
            \begin{multline}
                \nabla^2 J_\text{err} = 
                    G(\delta q)^T \! \left(
                    D(\delta q)^T Q D(\delta q) + \nabla D \right) G(\delta q)  \\
                    + I_3 (\phi^T Q  D(\delta q)) \delta q %\\
            \end{multline}
            where, for the Cayley map,
            \begin{equation}
                D(q) = \pdv{\varphi^{-1}}{q} = -\frac{1}{q_s^2}\begin{bmatrix}
                    q_v \;\; & -\frac{1}{q_s} I_3
                \end{bmatrix}
            \end{equation}
            \begin{equation}
                \nabla D = \pdv{q}(D(q)^T Q \phi) 
                = -\frac{1}{q_s^2} \begin{bmatrix} 
                    -2 \frac{q_v}{q_s}^T Q \phi & \phi^T Q \\
                                   Q \phi & 0 \\
                \end{bmatrix}.
            \end{equation}

@@

\section{Experiments} \label{sec:experiments}
    Code for all experiments is available on 
    \href{https://github.com/RoboticExplorationLab/PlanningWithAttitude}
    {GitHub\footnote{\url{https://github.com/RoboticExplorationLab/PlanningWithAttitude}}}.

@@

        \begin{figure}
            \centering
            \includegraphics[height=4cm, width=8cm]{figures/timing_chart.tikz}
            \caption{Timing results for the airplane barrell roll, quadrotor
                flip, and quadrotor zig-zag examples using iLQR with
                roll-pitch-yaw Euler angles (RPY), quaternions, and iMLQR. The
                timing result for the Quadrotor flip with Euler angles is omitted
                since it failed to converge.
            }
            \label{fig:timing_chart}
        \end{figure}

        \subsubsection{Quadrotor Zig-Zag}
            \begin{figure}[t]
                \centering
                \includegraphics[height=5.2cm]{figures/zig_zag_tracking.png}
                \caption{Zig-zag trajectory. The black quadrotors indicate
                    the reference trajectory generated by iMLQR. The yellow
                    quadrotors track the reference trajectory using time-varying
                    MLQR, while the red uses the HFCA controller from
                    \cite{watterson2020control}.
                }
                \label{fig:zig-zag}
            \end{figure}    
            \begin{figure}[t]
                \centering
                \includegraphics[height=5.2cm,trim={0 3cm 0 0},clip]{figures/quad_flip.png}
                \caption{Snapshots of the quadrotor flip trajectory. The
                    red-colored quadrotors represent the state near t=0 s and the
                    green-colored quadrotors represent the state near t=5.0 s
                }
                \label{fig:quad_flip}
            \end{figure}
